Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-21 paper/paper.tex
Section 4, minor

\subsection{Sorting Algorithms}\label{sec:sorting}
Our first case study is divide-and-conquer sorting algorithms encoded as
recursive hylomorphisms. Our accompanying code contains both mergesort and
quicksort as examples, but we will focus on quicksort for conciseness. We will
also show how to use program calculation to apply a fusion optimisation.  The
recursive structure of both mergesort and quicksort is a binary tree. For
example, in the case of quicksort, a list is split into a pivot, the label of
the node, and two sublists. We define the data functor of trees as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
To define an equivalent container, we notice that this functor has two shapes,
one for leaves, and one for nodes, and that nodes have two positions (akin to
\coq{l r : X} in \coq{i_node}). We also note that, in this container, positions
are only valid in nodes.
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
We name \coq{TreeF} the container that has these shapes and (base) poisitions.
For convenience, we define a series of wrappers to construct these containers:
% \mpav{The following algebra should be called a\_in or l\_in whatever}
\begin{coqcode}
Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
Notation a_node x l r := (MkCont (Node _ x)
  (fun p => match val p with | Lbranch => l | Rbranch => r end)).
\end{coqcode}
The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with the
following algebra and coalgebra. In the below definition, we use \coq{posL} and
\coq{posR} as wrappers for \coq{Lbranch} and \coq{Rbranch} with validity proofs.
\begin{coqcode}
Definition merge_f (x : App (TreeF unit int) (list int)) : list int :=
  let (sx, kx) := x in
  match sx return (Container.Pos sx -> _) -> _ with
  | Leaf _ _ => fun _ => nil
  | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
  end kx.
Definition qsplit_f (x : list int) : App (TreeF unit int) (list int) :=
  match x with
  | nil => a_leaf tt
  | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in a_node h l r
  end.
\end{coqcode}
Our tactics automatically discharge the proof that these are proper morphisms
(called \coq{merge} and \coq{qsplit}), and we prove that the coalgebra
\coq{qsplit} is recursive by showing that it respects the ``less-than'' relation
on the length of the lists.
%\begin{coqcode}
%Lemma split_fin : respects_relation split (@length int) lt.  (* *) Qed.
%
%Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
%  := mk_wf_coalg wf_lt split_fin.
%\end{coqcode}
The code that we extract for \coq{hylo merge qsplit} is similar to a hand-written implementation:
\begin{ocamlcode}
let rec qsort = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
    let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
    app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}

>>>>

\subsection{Sorting Algorithms}\label{sec:sorting}
We now focus on divide-and-conquer sorting algorithms and demonstrate how to use
program calculation to apply a fusion optimisation. Our accompanying code
contains both mergesort and quicksort as examples, but we will focus on
quicksort for conciseness.

The recursive structure of quicksort and mergesort is given by the following
functor
\begin{coqcode}
Inductive ITreeF N X := i_leaf | i_node (n : N) (l r : X)
\end{coqcode}
To explain this consider quicksort.
The list is split into a pivot which is represented by the and two sublists. We define the data functor of trees as follows:

To define an equivalent container, we notice that this functor has two shapes,
one for leaves, and one for nodes, and that nodes have two positions (akin to
\coq{l r : X} in \coq{i_node}). We also note that, in this container, positions
are only valid in nodes.
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
We name \coq{TreeF} the container that has these shapes and (base) positions.
For convenience, we define a series of wrappers to construct these containers:
% \mpav{The following algebra should be called a\_in or l\_in whatever}
\begin{coqcode}
Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
Notation a_node x l r := (MkCont (Node _ x)
  (fun p => match val p with | Lbranch => l | Rbranch => r end)).
\end{coqcode}
The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with the
following algebra and coalgebra. In the below definition, we use \coq{posL} and
\coq{posR} as wrappers for \coq{Lbranch} and \coq{Rbranch} with validity proofs.
\begin{coqcode}
Definition merge_f (x : App (TreeF unit int) (list int)) : list int :=
  let (sx, kx) := x in
  match sx return (Container.Pos sx -> _) -> _ with
  | Leaf _ _ => fun _ => nil
  | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
  end kx.
Definition qsplit_f (x : list int) : App (TreeF unit int) (list int) :=
  match x with
  | nil => a_leaf tt
  | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in a_node h l r
  end.
\end{coqcode}
Our tactics automatically discharge the proof that these are proper morphisms
(called \coq{merge} and \coq{qsplit}), and we prove that the coalgebra
\coq{qsplit} is recursive by showing that it respects the ``less-than'' relation
on the length of the lists.
%\begin{coqcode}
%Lemma split_fin : respects_relation split (@length int) lt.  (* *) Qed.
%
%Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
%  := mk_wf_coalg wf_lt split_fin.
%\end{coqcode}
The code that we extract for \coq{hylo merge qsplit} is similar to a hand-written implementation:
\begin{ocamlcode}
let rec qsort = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
    let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
    app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
